Parameter-less Population Pyramid

### 2014-03-22 Parameterless Population Pyramid/main.tex
Algorithms in, diagram out.  Fits to 8 pages!

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many GA techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid-like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution. Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optimum using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to perform crossover,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.

\begin{figure}
  \begin{algorithmic}
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber (Section~\ref{sec-hillclimber})
    \If{solution $\notin$ pyramid}
      \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
      \If{solution fitness has improved}
        \If{solution $\notin$ pyramid}
          \State Add solution to $P_{i+1}$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization.}
  \label{fig:psuedo}
\end{figure}

\begin{figure}
  \centering
  \includegraphicsfit{P3_big_fail}
  \caption{Example of the first steps of P3's optimization described in Section~\ref{sec-p3}.
  \#7 is not added because it is a duplicate of a solution already in the pyramid.  \#13 is
  not added because it is not a fitness improvement over \#12.}
  \label{fig-p3}
\end{figure}

@@

Along with storing the solutions, each level also maintains a table of pairwise
gene value frequencies for use by crossover.  This is also built constructively, such that initially
all frequencies are set to 0.  Each time a solution is added to a level, all
pairs of loci are queried for their gene values.  The table is then updated,
incrementing the count for each observed pairing.  This requires $O(N^2)$ operations
(no evaluations) each time a solution is added to a level of the pyramid, but is
constant in the number of solutions currently in the level.

@@

\begin{figure}
  \begin{algorithmic}
  \Procedure{Cluster-Creation}{}
    \State $unmerged \leftarrow \{\{i\} | 0 \le i < N\}$
    \State $useful \leftarrow unmerged$
    \While{$|unmerged|>0$}
      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
      \State $useful \leftarrow \{C_i \cup C_j\}$
      \If{$D(C_i, C_j) == 0$}
        \State $useful \leftarrow useful - \{C_i, C_j\}$
      \EndIf
    \EndWhile
    \State Order $useful$ based on cluster size, smallest first.
    \State Remove largest cluster from $useful$.

    \Return $useful$
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}.}
  \label{fig:cluster-creation}
\end{figure}

@@

Starting from the leaves, the binary tree is constructed agglomeratively. Each
iteration the clusters $C_i$ and $C_j$ which minimize $D(C_i, C_j)$ are
merged to form a new cluster $C_i \cup C_j$.
The process ends when there is only a single cluster
containing all loci.  This clustering algorithm
requires $O(N^2)$ operations to construct the linkage tree
from the frequency table~\cite{gronau:2007:upgma}.

Not all of  the resulting clusters in the linkage  tree are useful for
crossover.  The most obvious useless  cluster is one that contains all
the loci.   A crossover that preserves  all loci in  a solution cannot
create  a  new  individual.   We  further  remove  clusters  that  are
completely  contained   within their direct superset,   such  that  if
$D(C_i,C_j) = 0$, $C_i$  and  $C_j$ are  not used  when  performing
crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
perfectly predict the gene values in $C_j$ and vice versa. From this we can
conclude  that  the relative settings  for  those  loci  should  never  be
disrupted.  When  used on particularly small populations  this has the
potential  to  greatly  reduce  the  number of  clusters  used  during
crossover.  For  instance, if  $|P_i| = 1$,  no clusters exist. If
$|P_i| =  2$ at most two clusters  exist: one for all  genes where the
solutions agree and one for where they disagree.
This means that Figure~\ref{fig-p3} is infeasible as crossover
cannot make \#5 and \#9 unique from \{\#2, \#4\} and \{\#5, \#8\}, respectively.
This inaccuracy was kept to make Figure~\ref{fig-p3} better express
higher level concepts.

P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution as follows.  For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in that cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through fitness plateaus.

>>>>

\section{The P3 Algorithm}
\label{sec-p3}
Unlike many GA techniques, P3
does not have a single population of solutions.  Instead P3 maintains
a pyramid-like structure, such that each level of the pyramid has a
population of solutions, with each level representing something akin
to different generations of evolution.  Figure~\ref{fig-p3} provides
a high level summary of how P3 iteratively constructs these populations
and utilizes them to improve randomly generated solutions through
hill climbing and crossover.  While P3 can be applied to any discrete value
representation, we will focus on the binary domain here for simplicity.
\\
\begin{comment}
Consider the example evolution
presented in Figure~\ref{fig-p3}.  At the start of optimization P3
has no information about the landscape so solution \#1 is randomly generated.  Solution \#1
is brought to a local optimum using a hill climber
(Section~\ref{sec-hillclimber}) creating solution \#2, which is added
to the population pyramid (Section~\ref{sec-pyramid}) at population
level $P_0$.  With a population of 1 at $P_0$, there is insufficient information to perform crossover,
so solution \#3 is randomly generated and optimized via hill climbing
to create solution \#4, which is likewise added to the population
pyramid at population level $P_0$.
Crossover (Section~\ref{sec-crossover}) is applied to solutions \#2
and \#4 creating solution \#5. As \#5 used
information from $P_0$ it is added to the next level of the pyramid at
population level $P_1$.  Because of the crossover technique that P3
uses, there is no more information to be gained from the present
population. Thus, \#6 is randomly
generated and improved into \#7.  Unlike \#2 and \#4, \#7 is not added
in this example because it happens to be identical to a solution
already in the pyramid.  Even though \#7 is not unique, it still
crosses with $P_0$, which does make a unique improvement \#8 that is
added to $P_1$ and mixed with that level to create \#9.  As before,
the process creates \#10, \#11, and \#12 using random generation, hill
climbing, and crossover, respectively.  In this example \#13 is not
added to $P_2$ because it is not a strict fitness improvement over
\#12.  There is not enough information in $P_2$ to improve \#13
further, so the process would begin yet again with a random solution.
\end{comment}

\begin{figure}
  \begin{algorithmic}
  \Procedure{Iterate-P3}{}
    \State Create random solution
    \State Apply hill climber (Section~\ref{sec-hillclimber})
    \If{$\forall_{P_i\in pyramid}$ solution $\notin P_i$}
      \State Add solution to $P_0$ (Section~\ref{sec-pyramid})
    \EndIf

    \ForAll{$P_i \in pyramid$}
      \State Mix solution with $P_i$ (Section~\ref{sec-crossover})
      \If{solution's fitness has improved}
        \If{$\forall_{P_i\in pyramid}$ solution $\notin P_i$}
          \State Add solution to $P_{i+1}$
        \EndIf
      \EndIf
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{One iteration of P3 optimization. $pyramid$ is an
           ordered set of populations.}
  \label{fig-p3}
\end{figure}

\begin{comment}
\begin{figure}
  \centering
  \includegraphicsfit{P3_big_fail}
  \caption{Example of the first steps of P3's optimization described in Section~\ref{sec-p3}.
  \#7 is not added because it is a duplicate of a solution already in the pyramid.  \#13 is
  not added because it is not a fitness improvement over \#12.}
  \label{fig-p3}
\end{figure}
\end{comment}

@@

Along with storing the solutions, each level also maintains a table of pairwise
gene value frequencies for use by crossover.  This is also built constructively, such that initially
all frequencies are set to 0.  Each time a solution is added to a level, all
pairs of loci are queried for their gene values.  The table is then updated,
incrementing the count for each observed pairing.  This requires $O(N^2)$ operations
(no evaluations) each time a solution is added to a level of the pyramid, but is
constant in the number of solutions currently in the level.
\\

@@

\begin{figure}
  \begin{algorithmic}
  \Procedure{Cluster-Creation}{}
    \State $unmerged \leftarrow \{\{0\}, \{1\}, \{2\}, \dots, \{N-1\}\}$
    \State $useful \leftarrow unmerged$
    \While{$|unmerged|>1$}
      \State $C_i, C_j \leftarrow \min_{C_i,C_j \in unmerged} D(C_i, C_j)$
      \State $unmerged \leftarrow unmerged - \{C_i, C_j\} + \{C_i \cup C_j\}$
      \State $useful \leftarrow useful + \{C_i \cup C_j\}$
      \If{$D(C_i, C_j) = 0$}
        \State $useful \leftarrow useful - \{C_i, C_j\}$
      \EndIf
    \EndWhile
    \State Order $useful$ based on cluster size, smallest first.
    \State Remove largest cluster from $useful$.

    \Return $useful$
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm describing how clusters are created using Equation~\ref{eq-distance}
           for each $P_i$. $unmerged$ and $useful$ are ordered sets of sets of gene loci.}
  \label{fig-cluster-creation}
\end{figure}

@@

Figure~\ref{fig-cluster-creation} depicts the process of agglomeratively constructing
the linkage tree.  Starting with each gene location in separate clusters, Equation~\ref{eq-entropy}
is used to find which $C_i$ and $C_j$ in $unmerged$, the set of gene clusters, have the tightest
linkage (minimum distance).  These are merged to form a new cluster $C_i \cup C_j$, and
the separate pieces are removed from $unmerged$ while $useful$ tracks all clusters ever
created in the hierarchy.  This process iterates until all locations have been merged
into a single cluster.  With proper bookkeeping this algorithm will construct
the linkage tree from the frequency table in $O(N^2)$~\cite{gronau:2007:upgma}.

\begin{comment}
Starting from the leaves, the binary tree is constructed agglomeratively. Each
iteration the clusters $C_i$ and $C_j$ which minimize $D(C_i, C_j)$ are
merged to form a new cluster $C_i \cup C_j$.
The process ends when there is only a single cluster
containing all loci.  This clustering algorithm
requires $O(N^2)$ operations to construct the linkage tree
from the frequency table~\cite{gronau:2007:upgma}.
\end{comment}

Not all of  the resulting clusters in the linkage  tree are useful for
crossover.  The most obvious useless  cluster is one that contains all
the loci.   A crossover that preserves  all loci in  a solution cannot
create  a  new  individual.   We  further  remove  clusters  that  are
completely  contained   within their direct superset,   such  that  if
$D(C_i,C_j) = 0$, $C_i$  and  $C_j$ are  not used  when  performing
crossover.   $D(C_i,C_j)$ only  returns  $0$ when  the  loci in  $C_i$
perfectly predict the gene values in $C_j$ and vice versa. From this we can
conclude  that  the relative settings  for  those  loci  should  never  be
disrupted.  When  used on particularly small populations  this has the
potential  to  greatly  reduce  the  number of  clusters  used  during
crossover.  For  instance, if  $|P_i| = 1$,  no clusters exist. If
$|P_i| =  2$ at most two clusters  exist: one for all  genes where the
solutions agree and one for where they disagree.
\begin{comment}
This means that Figure~\ref{fig-p3} is infeasible as crossover
cannot make \#5 and \#9 unique from \{\#2, \#4\} and \{\#5, \#8\}, respectively.
This inaccuracy was kept to make Figure~\ref{fig-p3} better express
higher level concepts.
\end{comment}

\begin{figure}
  \begin{algorithmic}
  \Procedure{Cluster-Usage}{}
    \ForAll{$C_i \in useful$}
      \ForAll{$d \in shuffled(P_i)$}
        \State Copy $d$'s gene values for $C_i$ into solution
        \If{solution was changed}
          \If{solution's fitness decreased}
            \State Revert changes
          \EndIf
          \State \textbf{break}
        \EndIf
      \EndFor
    \EndFor
  \EndProcedure
\end{algorithmic}
  \caption{Algorithm describing how clusters are used to perform crossover.}
  \label{fig-cluster-usage}
\end{figure}

P3 uses crossover to recombine known information in a level of the
pyramid with a candidate solution as depicted in Figure~\ref{fig-cluster-usage}.
For each cluster $C_i$ found by \Call{Cluster-Creation}{}, choose a random
donor $d$ from the current level such that $d$'s gene values for at
least one gene in $C_i$ differs from the candidate solution.
Copy all of $d$'s gene values into the solution and evaluate the new
solution.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through fitness plateaus.

\begin{comment}
For
each cluster select a random donor from the population.  Copy the
donor's gene values for all loci in that cluster into the candidate
solution.  Evaluate the candidate.  If the fitness of the candidate
has been reduced, revert the candidate's gene values.  It is important
to note that candidate changes are kept in the case when the fitness
is \emph{unaffected}.  This allows crossover to drift, potentially
moving through fitness plateaus.
\end{comment}
